Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (*Sida fallax*)

### 2023-12-13 ms/ms.qmd
preparing final submission

\noindent \textbf{Premise of the study}  The adaptive significance of stomata on both upper and lower leaf surfaces, called amphistomy, is unresolved. A widespread association between amphistomy and open, sunny habitats suggests the adaptive benefit of amphistomy may be greatest in these contexts, but this hypothesis has not been tested experimentally. Understanding why amphistomy evolves can inform  its potential as a target for crop improvement and paleoenvironment reconstruction.

\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the log-ratio of photosynthesis in an amphistomatous leaf to that of the same leaf but with gas exchange blocked through the upper (adaxial) surface, which we term "pseudohypostomy". We used humidity to modulate stomatal conductance and thus compare photosynthetic rates at the same total stomatal conductance. We estimated $\mathrm{AA}$ and related physiological and anatomical traits in 12 populations, six coastal (open, sunny) and six montane (closed, shaded), of the indigenous Hawaiian species ʻilima (*Sida fallax*).

\textbf{Key results}  Coastal ʻilima leaves benefit `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy compared to their montane counterparts. Our evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because 1) they are thicker and therefore have lower CO$_2$ conductance through the internal airspace, and 2) that they benefit more because they have similar conductance on each surface, as opposed to most of the conductance being on the lower (abaxial) surface.

\textbf{Conclusions}  This is the first direct experimental evidence that amphistomy *per se* increases photosynthesis, consistent with the hypothesis that parallel pathways through upper and lower mesophyll increase the supply of CO$_2$ to chloroplasts. The prevalence of amphistomatous leaves in open, sunny habitats can partially be explained the increased benefit of amphistomy in `sun' leaves, but the mechanistic basis of this observation is an area for future research.

**Keywords:** amphstomy, leaf, light, Hawaiʻi, *Sida fallax*, stomata

@@

Here we focus on variation within an indigenous plant species *Sida fallax* Walp. (Malvaceae), known in the Hawaiian language as ʻilima. ʻIlima is found from sea level to elevations $>1000$ mas on multiple Hawaiian islands. Coastal populations are morphologically different from montane populations (\autoref{fig:study-system}). Coastal regions of Hawaiʻi are characterized by high sun exposure, warmer temperatures, high winds, salinity, and variation in water availability. Coastal populations of ʻilima tend to be short and prostrate which likely helps them to withstand the windy environment (\autoref{fig:study-system}a). The leaves of these populations are covered on both surfaces in dense, soft hairs that give the leaves a silvery green appearance (\autoref{fig:study-system}b), which helps mitigate water loss by reflecting solar radiation, thereby lowering leaf temperature [@ehleringer_pubescence_1978]. Montane regions, on the other hand, provide very different challenges. Many other tall species grow on the slopes of these wet mountainous regions, which makes light competition a factor that plants may need to adapt to. Possibly due to this, montane populations are erect and shrub- or tree-like, capable of growing meters tall with strong, woody stems. These individuals have smooth, green foliage with serrated edges. Montane populations exhibit traits that may help them to compete for light availability. This montane morphology is not found in *S. fallax* populations on other Pacific Islands [@pejhanmehr_phylogenetic_2022].

@@

\begin{figure}[H]
  \includegraphics[height=3.25in]{../figures/study-system.pdf}
  \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
  \caption{A. Typical growth form of montane (left) and coastal (right) ‘ilima plants and B. leaves. C. Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island). D. Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study. Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error. The habitats differ significantly in mean annual precipitation (top-left), solar radiation (bottom-left), temperature (top-right), and vegetation height (bottom-right). MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
  \label{fig:study-system}
\end{figure}

@@

At each site, we selected one representative leaf from one plant near the middle of the transect for gas exchange measurements using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, NE, USA). We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at saturating light ($\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), ambient CO$_2$ (415 ppm), and $T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}$. The midday irradiance in coastal ʻilima typically meets or even exceeds a PPFD of $2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$ and previous experiments with sun leaves revealed that $2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$ is always at or near saturating irradiance. Even though lower irradiance may be saturating for montane leaves, we used this higher value for all leaves to standardize conditions.

@@

\autoref{fig:ags-curve} illustrates our method for collecting $A \textendash g_\text{sw}$ curves. We collected two curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve. To control for order effects, we alternated between starting with amphi or pseudohypo leaf measurements, though we did not detect an effect of treatment order on $\mathrm{AA}$ (results not shown). In the field, we acclimated the focal leaf to saturating light and high relative humidity ($\mathrm{RH} = 70\%$), as described above, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation. Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity. All other environmental conditions in the leaf chamber remained the same. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process of acclimating the leaf to 70\% $\mathrm{RH}$ and inducing stomatal closure with low $\mathrm{RH}$ with the other treatment (amphi or pseudohypo).

@@

In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e. the curves overlapped for a range of $g_\text{sw}$ values). In those cases, we estimated $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap between the curves. In 2 of 12 leaves, the $A \textendash g_\text{sw}$ curves did not quite overlap because the minimum $g_\text{sw}$ of the amphi curve was slightly greater than the maximum $g_\text{sw}$ of the pseudohypo curve. In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves. The vertical lines in \autoref{fig:licor} show the $g_\text{sw}$ for each leaf. We estimated $\mathrm{AA}$ from $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ for each leaf using the log-response ratio shown above.

@@

<!-- \begin{align*} -->

<!--   b = & D_\text{wv} / v \\ -->

<!--   m = & \frac{\pi c ^ 2}{j ^ {0.5} (4 h j + \pi c)} -->

<!-- \end{align*} -->

<!-- $b$ is the diffusion coefficient of water vapor in air ($D_\text{wv}$) divided by the kinematic viscosity of dry air ($v$). $D_\text{wv} = 2.49 \times 10^{-5} ~ \text{m}^2 ~ \text{s}^{-1}$ and $v = 2.24 \times 10 ^ {-2} ~ \text{m}^3 ~ \text{mol} ^ {-1}$ at $25 ^ \circ$ [@monteith_principles_2013]. For kidney-shaped guard cells, $c = h = j = 0.5$. -->

@@

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants more than that of montane plants. $\mathrm{AA}$ was significantly greater than 0 (95% confidence intervals did not overlap 0) in 5 of 6 coastal leaves, but only 1 of 6 montane leaves (\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual curves). Overall, the average $\mathrm{AA}$ among coastal and montane leaves is $`r AA_coastal`$ and $`r AA_montane`$, respectively; the difference in average $\mathrm{AA}$ between habitat types is $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = `r AA_diff`$. Posterior predictions closely match observed values of $A$ (\autoref{fig:pp-licor}), indicating an adequate model fit from which we can interpolate between measurements reliably. It also suggests that slight extrapolation beyond the data should be reliable, but this is less certain. When we remove two leaves where we extrapolated slightly beyond fitted $A \textendash g_\mathrm{sw}$ curves, we estimate that $\mathrm{AA}_\text{coastal}$ is still positive, $`r AA_coastal1`$, but the difference between coastal and montane leaves is smaller,  $`r AA_diff1`$, and confidence intervals slightly overlap 0.  Maximum photosynthetic rate was slightly, but not significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['A']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['A']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['A']][['p.value']], 2)`$); total stomatal conductance was similar (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['gsw']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['gsw']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['gsw']][['p.value']], 2)`$) in coastal and montane leaves (\autoref{fig:habitat-Ags}). Water-use efficiency ($A / g_\mathrm{sw}$) was significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['wue']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['wue']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['wue']][['p.value']], 2)`$).

```{=tex}
\begin{figure}[H]
  \includegraphics{../figures/habitat-aa.pdf}
  \caption{Coastal leaves benefit more from amphistomy than montane leaves. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf. Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat. The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves. $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:habitat-aa}
\end{figure}
```

@@

Coastal and montane leaves have similar average $g_\mathrm{smax,ratio}$, the ratio of adaxial (upper) to total $g_\mathrm{smax}$, the anatomical maximum stomatal conductance to water vapor (\autoref{fig:habitat-gmaxratio}); coastal leaves have $`r gmaxratio_sitetype`$ higher $g_\mathrm{smax,ratio}$ than montane leaves, but the 95% confidence intervals overlap 0 difference. The $g_\mathrm{smax,ratio}$ is somewhat bimodal among sites. Some sites in both habitats have leaves with $g_\mathrm{smax,ratio} < 0.07$ and others with $g_\mathrm{smax,ratio} > 0.2$ (\autoref{fig:habitat-gmaxratio}). This is particularly noticeable in montane sites where those on the Big Island of Hawaiʻi all have low $g_\mathrm{smax,ratio}$ whereas those on Oʻahu have relatively high $g_\mathrm{smax,ratio}$. There is no relationship between $g_\mathrm{smax,ratio}$ and $\mathrm{AA}$ in either habitat (\autoref{fig:traits-aa}B; $\text{slope}~= `r gmaxratio_aa`$) in our sample.

```{=tex}
\begin{figure}[H]
  \includegraphics{../figures/traits-aa.pdf}
  \caption{Relationships between leaf amphistomy advantage, (A) leaf thickness and (B) $g_\mathrm{smax,ratio}$ among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value. Each leaf is from a different montane (blue) or coastal (orange) site. Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression. Symbols: $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:traits-aa}
\end{figure}
```

@@

The authors thank Kasey Barton and two anonymous reviewers for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, TM Perez for advice on leaf sectioning, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.

@@

Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo with a DOI and stable URL upon publication. Raw data will be deposited on Dryad with a DOI and stable URL upon publication. [THE GITHUB REPO AND DRYAD DATA ARE AVAILABLE TO REVIEWERS]

@@

>>>>

\noindent \textbf{Premise of the study}  The adaptive significance of stomata on both upper and lower leaf surfaces, called amphistomy, is unresolved. A widespread association between amphistomy and open, sunny habitats suggests the adaptive benefit of amphistomy may be greatest in these contexts, but this hypothesis has not been tested experimentally. Understanding amphistomy informs its potential as a target for crop improvement and paleoenvironment reconstruction.

\textbf{Methods}  We developed a method to quantify "amphistomy advantage", $\mathrm{AA}$, as the log-ratio of photosynthesis in an amphistomatous leaf to that of the same leaf but with gas exchange blocked through the upper surface (pseudohypostomy). Humidity modulated stomatal conductance and thus enabled comparing photosynthesis at the same total stomatal conductance. We estimated $\mathrm{AA}$ and leaf traits in six coastal (open, sunny) and six montane (closed, shaded) populations of the indigenous Hawaiian species ʻilima (*Sida fallax*).

\textbf{Key results}  Coastal ʻilima leaves benefit `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than montane leaves. Evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because 1) they are thicker and have lower CO$_2$ conductance through the internal airspace, and 2) that they benefit more because they have similar conductance on each surface, as opposed to most conductance being through the lower surface.

\textbf{Conclusions}  This is the first direct experimental evidence that amphistomy increases photosynthesis, consistent with the hypothesis that parallel pathways through upper and lower mesophyll increase CO$_2$ supply to chloroplasts. The prevalence of amphistomatous leaves in open, sunny habitats can partially be explained the increased benefit of amphistomy in `sun' leaves, but the mechanistic basis remains uncertain.

**Keywords:** amphistomy, Hawaiʻi, leaf, light, Malvaceae, photosynthesis, *Sida fallax*, stomata

@@

Here we focus on variation within an indigenous plant species *Sida fallax* Walp. (Malvaceae), known in the Hawaiian language as ʻilima. ʻIlima is found from sea level to elevations $>1000$ mas on multiple Hawaiian islands. Coastal populations are morphologically different from montane populations (\autoref{fig:study-system}). Coastal regions of Hawaiʻi are characterized by high sun exposure, warmer temperatures, high winds, salinity, and variation in water availability. Coastal populations of ʻilima tend to be short and prostrate which likely helps them to withstand the windy environment (\autoref{fig:study-system}a). The leaves of these populations are covered on both surfaces in dense, soft hairs that give the leaves a silvery green appearance (\autoref{fig:study-system}b), which helps mitigate water loss by reflecting solar radiation, thereby lowering leaf temperature [@ehleringer_pubescence_1978]. Montane regions, on the other hand, provide very different challenges. Many other tall species grow on the slopes of these wet mountainous regions, which makes light competition a factor that plants may need to adapt to. Possibly due to this, montane populations are erect and shrub- or tree-like, capable of growing meters tall with strong, woody stems. These individuals have smooth, green foliage with serrated edges. Montane populations exhibit traits that may help them to compete for light availability. This montane morphology is not found in *S. fallax* populations on other Pacific Islands [@pejhanmehr_low_2023].

@@

@@

At each site, we selected one representative leaf from one plant near the middle of the transect for gas exchange measurements using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA). We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at saturating light ($\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), ambient CO$_2$ (415 ppm), and $T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}$. The midday irradiance in coastal ʻilima typically meets or even exceeds a PPFD of $2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$ and previous experiments with sun leaves revealed that $2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$ is always at or near saturating irradiance. Even though lower irradiance may be saturating for montane leaves, we used this higher value for all leaves to standardize conditions.

@@

Appendix S1 (see the Supplementary Data with this article), \autoref{fig:ags-curve} illustrates our method for collecting $A \textendash g_\text{sw}$ curves. We collected two curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve. To control for order effects, we alternated between starting with amphi or pseudohypo leaf measurements, though we did not detect an effect of treatment order on $\mathrm{AA}$ (results not shown). In the field, we acclimated the focal leaf to saturating light and high relative humidity ($\mathrm{RH} = 70\%$), as described above, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation. Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity. All other environmental conditions in the leaf chamber remained the same. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process of acclimating the leaf to 70\% $\mathrm{RH}$ and inducing stomatal closure with low $\mathrm{RH}$ with the other treatment (amphi or pseudohypo).

@@

In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e. the curves overlapped for a range of $g_\text{sw}$ values). In those cases, we estimated $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap between the curves. In 2 of 12 leaves, the $A \textendash g_\text{sw}$ curves did not quite overlap because the minimum $g_\text{sw}$ of the amphi curve was slightly greater than the maximum $g_\text{sw}$ of the pseudohypo curve. In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves. The vertical lines in Appendix S1, \autoref{fig:licor} show the $g_\text{sw}$ for each leaf. We estimated $\mathrm{AA}$ from $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ for each leaf using the log-response ratio shown above.

@@

@@

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants more than that of montane plants. $\mathrm{AA}$ was significantly greater than 0 (95% confidence intervals did not overlap 0) in 5 of 6 coastal leaves, but only 1 of 6 montane leaves (\autoref{fig:habitat-aa}; see Appendix S1, \autoref{fig:licor} for individual curves). Overall, the average $\mathrm{AA}$ among coastal and montane leaves is $`r AA_coastal`$ and $`r AA_montane`$, respectively; the difference in average $\mathrm{AA}$ between habitat types is $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = `r AA_diff`$. Posterior predictions closely match observed values of $A$ (Appendix S1, \autoref{fig:pp-licor}), indicating an adequate model fit from which we can interpolate between measurements reliably. It also suggests that slight extrapolation beyond the data should be reliable, but this is less certain. When we remove two leaves where we extrapolated slightly beyond fitted $A \textendash g_\mathrm{sw}$ curves, we estimate that $\mathrm{AA}_\text{coastal}$ is still positive, $`r AA_coastal1`$, but the difference between coastal and montane leaves is smaller,  $`r AA_diff1`$, and confidence intervals slightly overlap 0.  Maximum photosynthetic rate was slightly, but not significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['A']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['A']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['A']][['p.value']], 2)`$); total stomatal conductance was similar (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['gsw']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['gsw']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['gsw']][['p.value']], 2)`$) in coastal and montane leaves (Appendix S1, \autoref{fig:habitat-Ags}). Water-use efficiency ($A / g_\mathrm{sw}$) was significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['wue']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['wue']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['wue']][['p.value']], 2)`$).

@@

Coastal and montane leaves have similar average $g_\mathrm{smax,ratio}$, the ratio of adaxial (upper) to total $g_\mathrm{smax}$, the anatomical maximum stomatal conductance to water vapor (Appendix S1, \autoref{fig:habitat-gmaxratio}); coastal leaves have $`r gmaxratio_sitetype`$ higher $g_\mathrm{smax,ratio}$ than montane leaves, but the 95% confidence intervals overlap 0 difference. The $g_\mathrm{smax,ratio}$ is somewhat bimodal among sites. Some sites in both habitats have leaves with $g_\mathrm{smax,ratio} < 0.07$ and others with $g_\mathrm{smax,ratio} > 0.2$ (Appendix S1, \autoref{fig:habitat-gmaxratio}). This is particularly noticeable in montane sites where those on the Big Island of Hawaiʻi all have low $g_\mathrm{smax,ratio}$ whereas those on Oʻahu have relatively high $g_\mathrm{smax,ratio}$. There is no relationship between $g_\mathrm{smax,ratio}$ and $\mathrm{AA}$ in either habitat (\autoref{fig:traits-aa}B; $\text{slope}~= `r gmaxratio_aa`$) in our sample.

@@

The authors thank Kasey Barton and three anonymous reviewers for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, TM Perez for advice on leaf sectioning, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB. This is publication \#213 from the School of Life Sciences, University of Hawaiʻi at Mānoa.

@@

Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-ilima) and archived on Zenodo (). Raw data are deposited on Dryad (https://doi.org/10.5061/dryad.rxwdbrvfw
).

@@

::: {#refs}
:::

\clearpage

# Figure Legends --- {#figure-legends}

\begin{figure}[H]
  % \includegraphics[height=3.25in]{../figures/study-system.pdf}
  % \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
  \caption{A. Typical growth form of montane (left) and coastal (right) ‘ilima plants and B. leaves. C. Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island). D. Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study. Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error. The habitats differ significantly in mean annual precipitation (top-left), solar radiation (bottom-left), temperature (top-right), and vegetation height (bottom-right). MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
  \label{fig:study-system}
\end{figure}

\begin{figure}[H]
  % \includegraphics{../figures/habitat-aa.pdf}
  \caption{Coastal leaves benefit more from amphistomy than montane leaves. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf. Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat. The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves. $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:habitat-aa}
\end{figure}

\begin{figure}[H]
  % \includegraphics{../figures/traits-aa.pdf}
  \caption{Relationships between leaf amphistomy advantage, (A) leaf thickness and (B) $g_\mathrm{smax,ratio}$ among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value. Each leaf is from a different montane (blue) or coastal (orange) site. Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression. Symbols: $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:traits-aa}
\end{figure}
```
